Variation in gaze understanding across the life span: A process-level perspective

### 2023-09-18 paper/manuscript.tex
complete first draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{social cognition, individual differences, gaze cues, cognitive modeling\newline\indent Word count: xxx}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Variation in gaze understanding across the life span: A process-level perspective},
  pdfauthor={Julia Prein1, Manuel Bohn1,2, Luke Maurits1, Annika Werwach1, \& Daniel B. M. Haun1},
  pdflang={en-EN},
  pdfkeywords={social cognition, individual differences, gaze cues, cognitive modeling},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Variation in gaze understanding across the life span: A process-level perspective}
\author{Julia Prein\textsuperscript{1}, Manuel Bohn\textsuperscript{1,2}, Luke Maurits\textsuperscript{1}, Annika Werwach\textsuperscript{1}, \& Daniel B. M. Haun\textsuperscript{1}}
\date{}

@@

The authors made the following contributions. Julia Prein: Conceptualization, Methodology, Software, Investigation, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Manuel Bohn: Conceptualization, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Luke Maurits: Formal Analysis, Writing - Review \& Editing; Annika Werwach: Methodology, Investigation, Writing - Review \& Editing; Daniel B. M. Haun: Supervision, Writing - Review \& Editing.

Correspondence concerning this article should be addressed to Julia Prein, Max Planck Institute for Evolutionary Anthropology, Deutscher Platz 6, 04103 Leipzig, Germany. E-mail: \href{mailto:julia_prein@eva.mpg.de}{\nolinkurl{julia\_prein@eva.mpg.de}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany\\\textsuperscript{2} Institute of Psychology, Leuphana University Lüneburg, Germany}

\abstract{%
Abstract
}

@@

@@

\begin{itemize}
\tightlist
\item
  why do we care about developmental trajectory? ref to stat learning paper
\item
  variation
\end{itemize}

\hypertarget{why-do-we-need-gaze-understanding}{%
\subsection{Why do we need gaze understanding?}\label{why-do-we-need-gaze-understanding}}

How do humans learn about their environment and navigate through their social surroundings?
One possibility to extract information from the environment is through following others' focus of attention.
Building a common ground is considered especially important in communicative interactions and shared activities (Tomasello, Hare, Lehmann, \& Call, 2007).

\hypertarget{how-does-gaze-following-emerge}{%
\subsection{How does gaze following emerge?}\label{how-does-gaze-following-emerge}}

Existing studies operationalize gaze following as the ability to follow another agent's line of sight.
As one of the most fundamental social-cognitive abilities, it has been extensively studied in infancy and early childhood.
Infants as young as six months can attune their gaze to that of another agent (D'Entremont, Hains, \& Muir, 1997).
At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access (Moll \& Tomasello, 2004).

While the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence.
One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy.However, many cognitive abilities develop with age (e.g., working memory, Gathercole, Pickering, Ambridge, \& Wearing, 2004).
Similarly, visual processing appears to improve with age.
Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill.

\hypertarget{the-scope-of-infants-gaze-following-ability}{%
\subsection{The scope of infants' gaze following ability}\label{the-scope-of-infants-gaze-following-ability}}

Though these studies suggest that young infants can align their visual attention to another's line of sight, it does not necessarily include understanding the intentions of the other agent.
Infants could simply attune their orientation or be attracted by others' gaze without processing what exactly the other is seeing (cf.~Butterworth \& Jarrett's ecological and geometric mechanism, Butterworth and Jarrett (1991){]}. Therefore, it is crucial to study children's intentional understanding of gaze.

Moore, Angelopoulos, and Bennett (1997) showed that 9-month-olds followed an agent's gaze more, when it was accompanied by a dynamic head turn in comparison to a static head turn.

In a hiding game with two search locations, Povinelli, Reaux, Bierschwale, Allain, and Simon (1997) found that three-year-olds used gaze as a cue to locate the reward, while two-year-olds performed at chance level.

In a similar object choice paradigm with two containers, Behne, Carpenter, and Tomasello (2005) investigated whether infants understand the communicative intent behind pointing and gaze cues.
In contrast to Povinelli et al. (1997), they found that already 14-month-olds used the agent's cues to select an object.
In conditions with absent-minded `cues', infants performed around chance.
This could be interpreted as infants recognizing the nature of this joint activity: namely, that the adult's behavior was beneficial and relevant for their object choice.

\hypertarget{head-vs-eye-direction}{%
\subsubsection{Head vs eye direction}\label{head-vs-eye-direction}}

It is important to note that in many existing gaze conditions, the experimenter shifted their eyes and head in synchrony (e.g., Behne et al. (2005)).
Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone.

A handful of studies approached this potential confound by separately manipulating head and eye movement.
Brooks and Meltzoff (2002) implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes.

Investigating the `cooperative eye hypothesis', Tomasello et al. (2007) implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither.
They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement.

Importantly, the subjects were not presented with an object choice but their attention orientation was measured.

\begin{itemize}
\item
  (Raviv \& Arnon, 2018)
\item
  (Astor \& Gredebäck, 2022)
\item
  (Colombo, 2001)
\item
  (Scaife \& Bruner, 1975)
\item
  (Itakura \& Tanaka, 1998)
\item
  (Carpenter, Nagell, \& Tomasello, 1998) ``Several other studies have attempted to determine more precisely the cue that infants are using when they follow the gaze direction of others, that is, whether they use adults' head or eye orientation. In tasks comparing infants' responses when the experimenters turned their head and eyes together to targets with their responses when the experimenters directed their eyes to the targets but their head remained facing forward, Corkum and Moore (1995), Lempers (1979), and Lempers, Flavell, and Flavell (1977) all found that only infants age 12 months and older responded correctly when eyes and head were oriented in the same direction and that infants at all ages (i.e., through 19 months) performed poorly when eye and head direction diverged'' (p.10-11) object choice.
\item
  (Silverstein, Feng, Westermann, Parise, \& Twomey, 2021) for vertical plane
\item
  (Zhang, Zhang, Zhang, Tang, \& Liu, 2019)
\item
  (Frischen, Bayliss, \& Tipper, 2007)
\item
  (Lee, Eskritt, Symons, \& Muir, 1998)
\item
  (Coelho, George, Conty, Hugueville, \& Tijus, 2006)
\end{itemize}

@@

\hypertarget{developmental-trajectory-measuring-modeling-individual-differences}{%
\subsubsection{Developmental trajectory, measuring \& modeling individual differences}\label{developmental-trajectory-measuring-modeling-individual-differences}}

In this study, we were interested in the developmental trajectory of gaze understanding.
While we expect the younger children to be able to follow gaze, we aimed at assessing the differentiation of their social-cognitive ability.
Our goal was \emph{not} to establish the youngest age at which children understand gaze cues.
Rather, we wanted to examine how that ability changes with age.

In our study, we focused on the communicative intents of gaze: we asked children to locate a target by following an agent's gaze.
While language demands were kept low, the participants had to actively respond and, therefore, make use of the presented gaze cue.

A unique contribution of this study is the richness of the data set.
Methodological challenges arise when trying to compare data across ages from qualitatively and quantitatively different study tasks.
We could circumvent these issues by applying the exact same task for the entire life span.

\hypertarget{lifespan}{%
\section{Lifespan}\label{lifespan}}

\begin{itemize}
\tightlist
\item
  development \& individual differences in gaze understanding
\item
  verweis methods paper reliable differences kinder \& adults.
\item
  kontinuerliche, systematische variation, wodrin? =\textgreater{} model
\end{itemize}

\url{https://osf.io/6yjz3}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:lifespan_sample}}

\begin{tabular}{lllll}
\toprule
Age group & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{Age mean} & \multicolumn{1}{c}{Age range} & \multicolumn{1}{c}{Age SD}\\
\midrule
3.00 & 19 (7 female) & 3.62 & 3.04 - 3.99 & 0.31\\
4.00 & 17 (9 female) & 4.45 & 4.05 - 4.91 & 0.30\\
5.00 & 22 (13 female) & 5.56 & 5.08 - 5.99 & 0.31\\
6.00 & 24 (16 female) & 6.50 & 6.1 - 6.99 & 0.28\\
7.00 & 39 (20 female) & 7.48 & 7.04 - 7.95 & 0.25\\
8.00 & 41 (20 female) & 8.46 & 8.03 - 8.98 & 0.27\\
9.00 & 56 (29 female) & 9.46 & 9.01 - 9.96 & 0.28\\
10.00 & 35 (22 female) & 10.49 & 10.01 - 11 & 0.28\\
11.00 & 54 (26 female) & 11.43 & 11.01 - 11.96 & 0.28\\
12.00 & 43 (19 female) & 12.41 & 12.01 - 12.99 & 0.30\\
13.00 & 42 (19 female) & 13.50 & 13.09 - 13.99 & 0.27\\
14.00 & 20 (14 female) & 14.37 & 14.05 - 14.98 & 0.23\\
15.00 & 21 (11 female) & 15.56 & 15.05 - 15.98 & 0.30\\
16.00 & 19 (10 female) & 16.51 & 16.17 - 16.97 & 0.24\\
17.00 & 19 (10 female) & 17.53 & 17.01 - 17.95 & 0.28\\
18.00 & 2 (0 female) & 18.00 & 18 - 18 & 0.00\\
19.00 & 5 (4 female) & 19.00 & 19 - 19 & 0.00\\
20.00 & 40 (25 female) & 23.02 & 20 - 29 & 2.77\\
30.00 & 40 (21 female) & 34.42 & 30 - 39 & 3.00\\
40.00 & 40 (24 female) & 44.17 & 40 - 49 & 2.92\\
50.00 & 40 (21 female) & 54.38 & 50 - 59 & 3.04\\
60.00 & 40 (21 female) & 63.73 & 60 - 69 & 2.56\\
70.00 & 40 (20 female) & 72.75 & 70 - 79 & 2.44\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

@@

We collected data from a remote child, teenager and adult sample.
For the remote child and teenager sample, we recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan\newline\indent Word count: xxx max 8000}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Variation in gaze understanding across the life span: A process-level perspective},
  pdflang={en-EN},
  pdfkeywords={social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Variation in gaze understanding across the life span: A process-level perspective}
\author{Julia Christin Prein\textsuperscript{1}, Luke Maurits\textsuperscript{1}, Annika Werwach\textsuperscript{1}, Daniel B. M. Haun\textsuperscript{1,*}, \& Manuel Bohn\textsuperscript{1,2,*}}
\date{}

@@

The authors made the following contributions. Julia Christin Prein: Conceptualization, Methodology, Software, Investigation, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Luke Maurits: Formal Analysis, Writing - Review \& Editing; Annika Werwach: Methodology, Investigation, Writing - Review \& Editing; Daniel B. M. Haun: Supervision, Writing - Review \& Editing; Manuel Bohn: Conceptualization, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing.

Correspondence concerning this article should be addressed to Julia Christin Prein, Max Planck Institute for Evolutionary Anthropology, Deutscher Platz 6, 04103 Leipzig, Germany. E-mail: \href{mailto:julia_prein@eva.mpg.de}{\nolinkurl{julia\_prein@eva.mpg.de}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany\\\textsuperscript{2} Institute of Psychology, Leuphana University Lüneburg, Germany\\\textsuperscript{*} Shared senior authorship}

\abstract{%
Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following in infancy has been thoroughly studied, we know little about the developmental trajectory throughout childhood and adolescence. In Study 1, we examine gaze understanding across the lifespan. In Study 2, we formalize the process of gaze understanding in a computational cognitive model that allows us to conceptualize individual differences in a psychologically meaningful way. The key parameter in our model is an inferential component, which describes how accurately a participant infers a target location based on an agent's gaze. In Study 3, we test a fundamental assumption of our gaze model by studying vector estimation in a non-social setting. Lastly, we explore the relationship between other social-cognitive abilities and gaze understanding. Our results suggest a steep learning curve in gaze understanding during the preschool years. In these years, children become more precise in locating the attentional focus of an agent. Precision levels then stay comparably stable, with a minor decay toward older adulthood. Our gaze model estimates reliable individual parameter values and recovers signature patterns in the data. Furthermore, we found that gaze understanding is associated with non-social vector estimation and perspective-taking but not with other Theory of Mind tasks. This work illustrates how the combination of reliable tasks and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
}

@@

\hypertarget{variation-in-gaze-understanding-across-the-life-span-a-process-level-perspective}{%
\section{Variation in gaze understanding across the life span: A process-level perspective}\label{variation-in-gaze-understanding-across-the-life-span-a-process-level-perspective}}

\textbf{Authors:} Julia Christin Prein\textsuperscript{1}, Luke Maurits\textsuperscript{1}, Annika Werwach\textsuperscript{1}, Daniel B. M. Haun\textsuperscript{1,*}, Manuel Bohn\textsuperscript{1,2,*}

\textbf{Affiliations:} \textsuperscript{1} Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany.
\textsuperscript{2} Institute of Psychology, Leuphana University Lüneburg, Germany.
\textsuperscript{*} shared senior authorship

\textbf{ORCiD}: \emph{Julia Christin Prein} \url{https://orcid.org/0000-0002-3154-6167}

\textbf{Conflicts of interest:} The authors declare that they have no conflict of interest.

\textbf{Data availability statement:} The web application (\url{https://ccp-odc.eva.mpg.de/tango-demo/}) described here is open source (\url{https://github.com/ccp-eva/tango-demo}). The data sets generated during and/or analysed during the current study are available in the {[}gazecues-modeling{]} repository (\url{https://github.com/ccp-eva/gazecues-modeling}). All experiments were pre-registered (\url{https://osf.io/zjhsc/}).

\textbf{Acknowledgements:} We thank Jana Jurkat for her help with data collection and participant recruitment. We would also like to thank Steven Kalinke for his technical programming support. We thank all the children, caregivers, and adults who participated in the study.

\textbf{Funding:} This study was funded by the Max Planck Society for the Advancement of Science, a noncommercial, publicly financed scientific organization (no grant number). Manuel Bohn was supported by a Jacobs Foundation Research Fellowship.

\textbf{Ethical statement:} The studies were approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany. Informed consent was obtained from all individual participants or their legal guardians.

\hypertarget{research-highlights-to-be-placed-before-the-abstract}{%
\section{Research highlights (to be placed before the abstract)}\label{research-highlights-to-be-placed-before-the-abstract}}

\begin{itemize}
\tightlist
\item
  Gaze understanding develops beyond infancy. Young adults reach the highest precision levels in localizing an attentional focus, while older adults perform slightly less precisely.
\item
  We present a computational model as a mechanistic description of gaze understanding. The model explains individual differences and recovers signature patterns in the data.
\item
  Our model assumes that gaze understanding relies on vector estimation. To test this assumption experimentally, we designed a non-social vector estimation task.
\item
  We found an association between gaze understanding and vector estimation, as well as Level 2 perspective-taking. Other Theory of Mind tasks did not correlate.
\end{itemize}

@@

How do humans learn about their environment and navigate through their social surroundings? A promising approach to extract information from the environment is through following the gaze of others. Recognizing the attentional focus of other agents also helps in building common ground and is vital for communicative interactions and shared activities (Tomasello, Hare, Lehmann, \& Call, 2007).

Existing studies operationalize gaze following as the ability to follow another agent's line of sight. As one of the most fundamental social-cognitive abilities, it has been extensively studied in infancy and early childhood. The traditional, often-used paradigm to study gaze following looks as follows (e.g., Gredebäck, Fikke, \& Melinder, 2010): The experimenter sits in front of the infant, with one object to their left and one to the right. The experimenter looks directly at the infant before shifting her head and eyes to one of the two objects. Infants' looking times or dichotomous object choices are then recorded. Analyses traditionally focus on the average age at which children reach an above-chance performance (i.e., looking at the gaze-cued target object). According to this research, infants as young as six months can attune their gaze to that of another agent (D'Entremont, Hains, \& Muir, 1997). At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access (Deák, Flom, \& Pick, 2000; Moll \& Tomasello, 2004).

At first glance, the debate about children's gaze-following abilities seems to be settled.

Some open questions remain, partly because of the methodological choices and the focus of previous research. For example, most studies focussed on infancy and focussed on the youngest age, in which children succeed. Often, these studies rely on two objects\ldots{}

On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants' abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further potential developments in childhood and adolescence; (2) Gaze cues are often accompanied by a head movement in the same direction, which makes it hard to isolate the signaling power of the eyes; (3) Often, infants' looking times are taken as a benchmark for passing a task, while it is uncertain how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Behavior is grouped into passing/failing or a comparison against chance, ignoring the accuracy of children; (6) Analyses focus on group-level averages, overlooking variation in the sample. In the following paragraphs, we will turn to each of these arguments.

While the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence. One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy. However, many cognitive abilities develop with age (e.g., working memory, Gathercole, Pickering, Ambridge, \& Wearing, 2004). Similarly, visual processing appears to improve with age. Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill. We believe that the development of gaze understanding is, most likely, not complete merely because infants can differentiate between two objects, one which has been gazed at. In daily life, we often face more nuanced, fine-grained gaze cues and must differentiate between a virtually unlimited number of potential targets.

In traditional gaze-following studies, the experimenter often shifts their eyes and head in synchrony (e.g., Behne, Carpenter, \& Tomasello, 2005). Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. Brooks and Meltzoff (2002) implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes. Moore, Angelopoulos, and Bennett (1997) showed that 9-month-olds followed an agent's gaze more when it was accompanied by a dynamic head turn in comparison to a static head turn. Investigating the `cooperative eye hypothesis', Tomasello et al. (2007) implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither. They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement. Results of Corkum and Moore (1995), Lempers (1979), and Lempers, Flavell, and Flavell (1977) suggest that infants, at least until 19 months, struggle when eye and head direction diverge. This underlines how important the movement of the head is when interpreting the gaze of others and that the scope of infants' gaze understanding might be less comprehensive than previously assumed.

Next, let us consider the classical response measure of collecting looking times. Though previous studies using the above-mentioned methods suggest that young infants can align their visual attention to another's line of sight, it does not necessarily include understanding the intentions or the content of the other agent's gaze. Infants could simply attune their orientation or be attracted by others' gaze without processing what exactly the other is seeing (cf.~Butterworth \& Jarrett's ecological and geometric mechanism, Butterworth \& Jarrett, 1991). Arguably, this is not sufficient to claim an understanding of visual perspectives, because the ``target'' or ``object of representation'' is not necessarily specified (Perner, Brandl, Garnham, \& Peter Lang, 2003, p. 358). To fulfill this criterion, Moll and Tomasello (2006) used an active behavioral choice as a response measure and tested infants' understanding of the intentions behind gaze. Infants interactively searched for the target object, which was either occluded or visually accessible. They found that 24-month-olds adaptively handed over the target object, while 18-month-olds still struggled to complete this task. In a hiding game with two search locations, Povinelli, Reaux, Bierschwale, Allain, and Simon (1997) found that two-year-olds performed at chance level when using gaze as a cue to locate the reward, while three-year-olds passed the task. However, in a similar object choice paradigm with two containers, Behne et al. (2005) investigated whether infants understand the communicative intent behind pointing and gaze cues. In contrast to Povinelli et al. (1997), they found that already 14-month-olds used the agent's cues to select an object. In conditions with absent-minded `cues', infants performed around chance. The specific scope of children's gaze understanding and whether looking toward the cued location translates into meaningful active behavior deserve further investigation.

Finally, presenting two objects, focusing on group-level analysis, and passing versus failing a task drastically limit the level of detail with which we can gauge children's gaze understanding. Undoubtedly, children can vary in the onset of their social-cognitive abilities: these skills develop over time, with experience, and on a continuum. Birch et al. (2017) argued that individuals can differ in their frequency (or propensity) and accuracy of applying social cognition in their daily lives. Measuring individual differences in this domain is especially important to understand the underlying processes and to disentangle the impact of environmental influences (scaffolding) and other cognitive abilities (Birch et al., 2017).

@@

With our study, we aim to deepen our knowledge about the development and mechanisms behind gaze understanding. We study the differentiation of this social-cognitive ability by using a study design with subtle gaze cues without head rotation. While we expect the younger children to be able to follow gaze, we aim at assessing the differentiation of their social-cognitive ability. Our goal was \emph{not} to establish the youngest age at which children understand gaze cues. Rather, we wanted to examine \emph{how} that ability changes with age. In our study, we focused on the communicative intents of gaze: we asked children to locate a target by following an agent's gaze. While language demands are low, participants have to actively respond and, therefore, make use of the presented gaze cue. By measuring a continuous, active behavior, we can capture meaningful variation between individuals. We strive to embrace and explain diversity in gaze understanding and use individual differences to inform what happens on a process level.

The present paper is structured in the following way. In Study 1, we explore gaze understanding across the lifespan. We assess the developmental trajectory and individual differences in this fundamental social-cognitive ability in a sample of 3- to 80-year-olds. In Study 2, we propose a computational cognitive model of gaze understanding. This model helps us to explain where individuals vary and which mechanisms participants use to locate an agent's attentional focus. In Study 3, we revisit a fundamental assumption of our computational gaze model by experimentally isolating the vector estimation component of gaze understanding. Furthermore, we investigate the relationship between gaze understanding and other social-cognitive processes.

\hypertarget{study-1-gaze-understanding-across-the-lifespan}{%
\section{Study 1: Gaze understanding across the lifespan}\label{study-1-gaze-understanding-across-the-lifespan}}

We aimed to assess the developmental trajectory of gaze understanding across the lifespan. First, we were interested in how this ability changes with age: Is the ability to understand gaze fully developed once emerged in infancy and stays stable across the lifetime? Or do we fine-tune our already existing ability in early adulthood? Do we then potentially even notice an age decay later in adulthood? Second, we were interested in individual differences across the age groups. Does the variation between individuals decrease, the older they get?
Do all adults reach the same level of precision in gaze understanding?

Due to the sheer number of participants, we restricted this analysis to a remote sample. Pre-registrations can be found here: \url{https://osf.io/snju6} (child sample) and \url{https://osf.io/6yjz3} (adult sample). The study obtained ethical clearance by the MPG Ethics commission Munich, Germany, falling under an umbrella ethics application (Appl. No.~2021\_45), and was approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology. The research adheres to the legal requirements of psychological research with children in Germany. Data was collected between May 2021 and April 2023.

@@

We collected data from a remote child, teenager and adult sample, reaching from 3 to 80 years of age (see Supplements for further details). The child and teenager sample consisted of 471 participants. We recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.
